Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
Sections 1-4: minor enhancements.

The Third Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13;
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized in July 2014 jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14; and
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized in July 2015 again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering, but WSSSPE2 didn't have a good way to enact those mechanisms, or to
encourage the attendees to follow through on their intentions.

The workshop included multiple mechanisms for participation and encouraged team
building around solutions. It strongly encouraged participation of early-career
scientists, postdoctoral researchers, and graduate students, with funds provided
to the conference organizers by the Moore Foundation, the National Science
Foundation (NSF), and the Software Sustainability Institute, to support the
travel of potential participants who would not otherwise be able to attend,
early-stage researchers, and those from underrepresented groups. These funds
allowed 16 additional participants to attend.

This report is based on collaborative notes taken with during the workshop,
which were linked from the GitHub issues that represented the potential and
actual working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}); the keynotes (\S\ref{sec:keynote}); a series of
lightning talks (\S\ref{sec:lightning}), intended to give an opportunity for
attendees to quickly highlight an important issue or a potential solution. The
report also gives summaries of action plans proposed by the working groups
(\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}). Lists of the
organizing committee (Appendix~\ref{sec:orgcom}), the registered attendees
(Appendix~\ref{sec:attendees}), and the travel award recipients
(Appendix~\ref{sec:awardees}) are provided. Finally, the report includes longer
descriptions of the activities that occurred in each of the working groups that
made substantial progress
(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).

@@

The first call for participation requested lightning talks, where authors could
make a brief statement about work that either had been done or was needed, with
the goal of contributing to the discussion of one or more working groups. There
was 24 lightning talks submitted, and, after a peer-review process, 16 of these
were accepted, as discussed further in Section~\ref{sec:lightning}.

The first call also discussed the potential action topics that came out of
WSSSPE2, and requested additional suggestions. The combination of existing and
new topics led to the following 18 potential topics that were advertised in the
call for participation:

@@

\item Development and Community
\begin{itemize}
\item Writing a white paper/review paper about best practices in developing
sustainable software
\item Documenting successful models for funding specialist expertise in software
collaborations
\item Creating and curating catalogs for software tools that aid sustainability
(perhaps categorized by domain, programming languages, architectures, and/or
functions, e.g., for code testing, documentation)
\item Documenting case studies for academia/industry interaction
\item Determining effective strategies for refactoring/improving legacy
scientific software
\item Determining principles for engineering design for sustainable software
\item Create a set of guidance giving examples of specific metrics for the
success of scientific software in use, why they were chosen, what they are
useful to measure, and any challenges/pitfalls; then publish this as a white
paper
\end{itemize}

@@

\item Publishing
\begin{itemize}
\item Developing a categorization of journals that publish software papers
(building on existing work), and case studies of alternative publishing
mechanisms that have been shown to improve software discoverability/reuse e.g.,
popular blogs/websites
\item Determining what journals that publish software paper should provide to
their reviewers (e.g., guidelines, mechanisms, metadata standards, etc.)
\end{itemize}

@@

WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
Department of Astronomy, University of Illinois titled \emph{Why Sustain
Scientific Software?}. Turk is a prolific scientific software practitioner and
has extensive experiences working on large collaborative projects employing
modern computing tools. He also co-organizes and champions WSSSPE events.

@@

WSSSPE3 left the traditional Supercomputing Conference environment this year,
and in Turk's words, this spoke to the fact that scientific software comes from
many different types of inquiries, deployment, strategies for maintenance,
users, and ways of measuring the value of a piece of software. It appeared to
Turk that the supercomputing community generally adopt some top-down approaches,
whereas the SciPy community more often than not uses more bottom-up systems. The
essential messages perceived were also often bipolar: the supercomputing
community thinks software is getting harder, with exascale computing and
optimization issues in mind, but the SciPy community thinks software becoming
better, with emerging tools such as Jupiter and productivity packages for
research workflow. Admitting such comparisons are somewhat unfair
generalizations, Turk reminded the audience that the different approaches bring
different types of ideas to the table, and he welcomed WSSSPE3 being conducted
outside existing preconceptions.

@@

At WSSSPE1, several models were presented for ensuring sustainability. Turk
considered that a familiar one was a funded piece of software where an external
agency provided funds to a group who are not necessarily exclusively working on
and developing a piece of software, keeps it going, and provides it to the
scientific community. The model of productized software, in which a piece of
software has grown into something that individuals rely on to the point that
research groups or people are willing to support it with some amount of funding,
for instance, a subscription to use cloud services that deploy a piece of
software, or purchase of a piece of software. A final model Turk felt conflicted
about is a volunteer model that is traditional old-school---not modern-day open
source---development.

Turk discussed whether productizing scientific software was synonymous with
being sustainable and self-sufficient. He thought it was not necessarily the
case and furthermore, it could lead to a divergence of interest between users
and developers.

Turk reminded the audience that the volunteer model means unpaid labor. On this
note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
the open source software
community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
Often times, a person funded to work full time on a scientific project can spend
a small amount of time for working on a piece of software necessary for that
project. However, all researchers' ability to participate in that volunteer
community are not always the same and may not always be aligned with their
research projects. From Turk's experience, we cannot always rely on unpaid labor
and volunteer time to sustain a piece of software---this came down to the
notions of the top-down and the bottom-up approaches, i.e., the funded versus
the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
projects can be just as large as a top-down software development project.

Turk said that sustaining scientific software really meant to him conducting
scientific inquiries, often by some specific software, and sustaining the people
we care about, our careers, and the future of our fields. According to Turk, we
all have an invested stake in sustaining scientific software. Hence, having
``sustained'' projects can suffocate new projects, so we need to make sure we
don't cause novel ideas and packages to suffer at the hands of the status quo.
%This is a straw man concept, but it's something to look out for, since
organizations will act in their own interest.

Turk talked about possible reasons why we want to sustain scientific software:
devotion to science and interests in pursuing the next stage of research; fun
and creative thrill in writing codes and papers; usefulness with measurable
impacts, for example, LINPACK and HDF groups providing of data storage to
satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
his wish-list of questions to be answered in the future:
%
\begin{itemize}

@@

\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
Facilitating Research Training While Supporting Scientific Research}} A case
study: ``The Cooperative Computing Lab'' at the University of Notre Dame is a
small group of individuals whose main tasks are collaborating with people that
have large-scale computing problems, operating various parallel computer
systems, conducting computer science research, and developing open source
software. One of the main challenges they face is finding the balance between
flexibility/training and stability/quality. Their current solution for ensuring
the latter was to add a software engineer (the presenter) to the existing team
of faculty and students, who now also serves as ``spring'' between flexibility
and stability.

>>>>

The Third Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13;
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14; and
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational; and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged a broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future for software development and usage in science and
engineering. But WSSSPE2 didn't have a good way to enact those mechanisms, or to
encourage the attendees to follow through on their intentions.

The WSSSPE3 workshop included multiple mechanisms for participation and encouraged team
building around solutions. It strongly encouraged participation of early-career
scientists, postdoctoral researchers, and graduate students, with funds provided
to the conference organizers by the Moore Foundation, the National Science
Foundation (NSF), and the Software Sustainability Institute, to support the
travel of potential participants who would not otherwise be able to attend,
early-stage researchers, and those from underrepresented groups. These funds
allowed 16 additional participants to attend.

This report is based on collaborative notes taken during the workshop,
which were linked from the GitHub issues that represented the potential and
actual working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
lightning talks (\S\ref{sec:lightning}). The
report also gives summaries of action plans proposed by the working groups
(\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}). Lists of the
organizing committee (Appendix~\ref{sec:orgcom}), the registered attendees
(Appendix~\ref{sec:attendees}), and the travel award recipients
(Appendix~\ref{sec:awardees}) are compiled. Finally, the report includes longer
descriptions of the activities that occurred in each of the working groups that
made substantial progress
(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).

@@

The first call for participation requested lightning talks, where each author could
make a brief statement about work that either had been done or was needed, with
the goal of contributing to the discussion of one or more working groups. There
were~24 lightning talks submitted, and, after a peer-review process, 16 of those
were accepted, as discussed further in Section~\ref{sec:lightning}.

The first call also discussed the potential action topics that came out of
WSSSPE2, and requested additional suggestions. The combination of existing and
new topics led to the following 18 potential topics that were advertised in the subsequent
calls for participation:

@@

\item Development and Community
\begin{itemize}
\item Writing a white paper/review paper about best practices in developing
sustainable software
\item Documenting successful models for funding specialist expertise in software
collaborations
\item Creating and curating catalogs for software tools that aid sustainability
(perhaps categorized by domain, programming languages, architectures, and/or
functions, e.g., for code testing, documentation)
\item Documenting case studies for academia/industry interaction
\item Determining effective strategies for refactoring/improving legacy
scientific software
\item Determining principles for engineering design for sustainable software
\item Create a set of guidance giving examples of specific metrics for the
success of scientific software in use, why they are chosen, what they are
useful to measure, and any challenges/pitfalls; then publish this as a white
paper
\end{itemize}

@@

\item Publishing
\begin{itemize}
\item Developing a categorization of journals that publish software papers
(building on existing work), and case studies of alternative publishing
mechanisms that have been shown to improve software discoverability/reuse, e.g.,
popular blogs/websites
\item Determining what journals that publish software paper should provide to
their reviewers (e.g., guidelines, mechanisms, metadata standards)
\end{itemize}

@@

WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
Department of Astronomy, University of Illinois, titled \emph{Why Sustain
Scientific Software?}. Turk is a prolific scientific software practitioner and
has extensive experiences working on large collaborative projects employing
modern computing tools. He also co-organizes and champions WSSSPE events.

@@

WSSSPE3 left the traditional Supercomputing Conference environment this year,
and in Turk's words, this spoke to the fact that scientific software comes from
many different types of inquiries, deployment, strategies for maintenance,
users, and ways of measuring the value of a piece of software. It appeared to
Turk that the supercomputing community generally adopts some top-down approaches,
whereas the SciPy community more often than not uses more bottom-up systems. The
essential messages perceived were also often bipolar: the supercomputing
community thinks that software is getting harder, with exascale computing and
optimization issues in mind; but the SciPy community thinks that software is becoming
better, with emerging tools such as Jupiter and productivity packages for
research workflows. Admitting such comparisons are somewhat unfair
generalizations, Turk reminded the audience that the different approaches bring
different types of ideas to the table, and he welcomed WSSSPE3 being conducted
outside existing preconceptions.

@@

At WSSSPE1, several models were presented for ensuring sustainability. Turk
considered that a familiar one was a funded piece of software where an external
agency provided funds to a group who are not necessarily exclusively working on
and developing the software, keeps it going, and provides it to the
scientific community. The model of productized software, in which a piece of
software has grown to the point that
research groups or people are willing to support it with some amount of funding,
for instance, a subscription to use cloud services that deploy a piece of
software, or purchase of a piece of software. A final model Turk felt conflicted
about is a volunteer model that is traditional old-school---not modern-day open
source---development.

Turk discussed whether productizing scientific software was synonymous with
being sustainable and self-sufficient. He thought it was not necessarily the
case and furthermore, it could lead to a divergence of interests between users
and developers.

Turk reminded the audience that the volunteer model means unpaid labor. On this
note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
the open source software
community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
Often times, a person funded to work full time on a scientific project can spend
a small amount of time for working on a piece of software necessary for that
project. However, researchers' abilities to participate in that volunteer
community are not always the same and may not always be aligned with their
research projects. From Turk's experience, we cannot always rely on unpaid labor
and volunteer time to sustain a piece of software---this came down to the
notions of the top-down and the bottom-up approaches, i.e., the funded versus
the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
projects can be just as large-scale as a top-down software development project.

Turk said that sustaining scientific software really meant to him conducting
scientific inquiries, often by some specific software, and sustaining the people
we care about, our careers, and the future of our fields. According to Turk, we
all have an invested stake in sustaining scientific software. Hence, having
``sustained'' projects can suffocate new projects, so we need to make sure we
don't cause novel ideas and packages to suffer at the hands of the status quo.

Turk talked about possible reasons why we want to sustain scientific software:
devotion to science and interests in pursuing the next stage of research; fun
and creative thrill in writing codes and papers; usefulness with measurable
impacts, for example, LINPACK and HDF groups providing data storage to
satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
his wish-list of questions to be answered in the future:
%
\begin{itemize}

@@

The lightning talks were intended to give an opportunity for
attendees to quickly highlight an important issue or a potential solution. 
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
Facilitating Research Training While Supporting Scientific Research}} A case
study: ``The Cooperative Computing Lab'' at the University of Notre Dame is a
small group of individuals whose main tasks are collaborating with people that
have large-scale computing problems, operating various parallel computer
systems, conducting computer science research, and developing open source
software. One of the main challenges they face is finding the balance between
flexibility/training and stability/quality. Their current solution for ensuring
the latter was to add a software engineer (the presenter) to the existing team
of faculty and students, who now also serves as ``spring'' between flexibility
and stability.
